### 2021-09-06 manuscript/methodology.tex
methodology until "Computational performance" - partial

Equation \ref{eq:integral-sum_mag} can be rewritten in matrix notation as follows:
\begin{equation}
\mathbf{d}(\mathbf{p}) = \mathbf{A} \mathbf{p} \: ,
\label{eq:predicted-data-vector_mag}
\end{equation}
where $\mathbf{d}(\mathbf{p})$ is the $N \times 1$ predicted data vector with $i$th element defined
be the predicted total-field anomaly $\Delta T_{i}$ (equation \ref{eq:integral-sum_mag}),
$\mathbf{p}$ is the $M \times 1$ parameter vector whose $j$th element is the magnetic moment intensity
$p_{j}$ of the $j$th dipole and $\mathbf{A}$ is the $N \times M$ sensitivity matrix with element 
$ij$ defined by the harmonic function $a_{ij}$ (equation \ref{eq:aij_mag}).
In the classical equivalent-layer technique, the common approach for 
estimating the parameter vector $\mathbf{p}$ from the observed 
total-field anomaly data $\mathbf{d}^{o}$ is solving the least-squares normal equations
\begin{equation}
\mathbf{A}^{\top}\mathbf{A} \: \mathbf{p} = 
\mathbf{A}^{\top} \mathbf{d}^{o} \: .
\label{eq:normal-equations}
\end{equation}
Equation \ref{eq:normal-equations} is usually solved by first computing the Cholesky 
factor of matrix $\mathbf{A}^{\top}\mathbf{A}$ and then using it to solve the linear 
system \citep[][ p. 262]{golub-vanloan2013}. 
This approach to estimate the parameter vector will be 
referenced throughout this work as the \textit{classical method}.

The computational cost required for estimating the classical solution can be very high
when dealing with large datasets. In the following subsections, we will show how to 
explore the structure of the sensitivity matrix $\mathbf{A}$ and 
efficiently solve the least-squares normal equations (equation \ref{eq:normal-equations}).

@@

Note that the standard CGLS solution (Algorithm \ref{al:std-cgls-algorithm}) requires 
neither inverse matrix nor matrix-matrix product. Instead, it only requires: one matrix-vector 
product out of the loop and two matrix-vector products per iteration (in steps 3 and 6). 
These products can be efficiently computed by using the 2D FFT, as a discrete convolution
(see Appendix A). \citet{takahashi2020convolutional} used this approach
to develop an efficient algorithm for gravity data processing. This modified approach in which
the standard CGLS method is modified to compute the matrix-vector products efficiently will be 
referenced throughout this work as the \textit{convolutional equivalent layer method}.

@@

%======================================================================================
\subsection{CGLS matrix-vector substitution}
%======================================================================================

Also differently for the symmetric sensitivity matrix described by \cite{takahashi2020convolutional}, the non-symmetric BTTB matrix cannot be reconstructed only by its first column. The construction of the matrix $\mathbf{A}$ (equation \ref{eq:aij_mag}) needs four columns: the first and last columns of the first column of blocks and the first and last columns of the last column of blocks. This has a physical implication in the equivalent layer which is not possible to use only one equivalent source to reprduce all elements of matrix $\mathbf{A}$, such as in the gravity case as demonstrared by \cite{takahashi2020convolutional}. Rather, in the magnetic case it takes four equivalent sources positioned at each corner of the equivalent layer. Figure \ref{fig:4_equivalent_sources} shows the positioning of the equivalent sources in a regular grid $N_x = 4$
and $N_y = 3$ necessary to calculate the four columns capable of recover the matrix $\mathbf{A}$.

In this work, we propose a different approach, by calculating the first column of all six different components of second derivatives matrices from $\mathbf{H}_{ij}$ (equation \ref{eq:Hi}). These matrices are, in fact, symmetrics or skew-symmetrics BTTBs, meaning that the first column has all elements of each matrix.

As pointed earlier in this work, the main improvement inside the CGLS method (Algorithm \ref{al:std-cgls-algorithm}) for estimating the parameter vector $\hat{\mathbf{p}}$ (equation \ref{eq:estimated-p-parameter-space}) is to substitute the matrix-vector multiplication $\mathbf{A}^{\top} \mathbf{s}^{(0)}$ out of the loop and the two matrix-vector multiplications inside the loop at steps 3 an 6, $\mathbf{A} \, \mathbf{c}^{(it)}$ and $\mathbf{A}^{\top} \, \mathbf{s}^{(it + 1)}$, that is necessary at each iteration and takes most of its runtime.

Our method consists in calculating the six first columns of the second derivatives of $\mathbf{H}$ (equation \ref{eq:Hi}) and embedding them into the first six columns of the block-circulant circulant-block (BCCB) matrices related to the $\mathbf{H}$ components. Thus, it is possible to calculate the first column of the BCCB matrix embbeded from matrix $\mathbf{A}$ (equation \ref{eq:aij_mag}) by multiplying each component with its respective constants and summing as shown in equation \ref{eq:aij_mag_expand}. In \cite{takahashi2020convolutional}, Appendix A, the authors demonstrated in details how to transform a symmetric BTTB matrix into a BCCB matrix $\mathbf{C}$. The process here is the same and that work can be referenced to achieve the same results.

To compare the efficiency of our algorithm we will use a numerical approach and calculate the floating-point operations (\emph{flops}), i.e., count the number of mathematical operations necessary to complete the estimative of parameter vector $\mathbf{\hat{p}}$ of the normal equations (equation \ref{eq:estimated-p-parameter-space}) and both the CGLS methods (algorithm \ref{al:std-cgls-algorithm}) for calculating the matrix-vector product by its standart way and our approach.

The \emph{flops} needed to solve the linear system in equation \ref{eq:estimated-p-parameter-space} using the Cholesky factorization is:
\begin{equation}
f_{classical} =  \dfrac{7}{3} N^{3} + 6 N^{2}\: ,
\label{eq:flops-normal-cholesky}
\end{equation}
where $N$ is the total number of observation points and also the size of estimated parameter vector $\mathbf{\hat{p}}$.

For the more efficient CGLS algorithm the estimative can be done in:
\begin{equation}
f_{cgls} =  2 N^{2} + it \, (4 N^{2} + 12 N) \: .
\label{eq:flops-cgls}
\end{equation}
However, our approach reduces further to:
\begin{equation}
f_{ours} =  \kappa  \, 16 N \log_2(4 N) + 24 N + it \, (\kappa  \, 16 N \log_2 (4 N) + 60 N) \: ,
\label{eq:flops-cgls-bccb}
\end{equation}
where $\kappa$ depends on the FFT algorithm. By default, in this work we will use $\kappa = 5$ for the \emph{radix-2} algorithm \citep{vanloan1992}.

Figure \ref{fig:flops} shows a comparative between the methods varying the number of observation points up to $1,000,000$, where it is possible to observe a reduction of $10^7$ orders of magnitude to estimate parameter vector $\mathbf{\hat{p}}$ in relation to the non-iterative classical method and $10^3$ orders of magnitude in relation to the standart CGLS algorithm using $50$ iterations. A more detailed, step by step, flops count of the classical and CGLS algorithm can be found in Appendix A.

@@

In Figure \ref{fig:sources_time} a comparison between the time to complete the task to calculate the first column of the BCCB matrix embbeded from the from $\mathbf{A}$ (equation \ref{eq:aij_mag}) by using only one equivalent source, i.e., calculating all six first column of the second derivatives matrices from $\mathbf{H}$ (equation \ref{eq:Hi}) and using four equivalent sources to calculate the four necessary columns from the non-symmetric matrix $\mathbf{A}$ (equation \ref{eq:aij_mag}). Although, very similar in time, with one source a small advantage can be observed as the number of data $N$ increases and goes beyond $N = 200,000$. This test was done from $N = 10,000$ to $N = 700,000$ with increases of $5,625$ observation points.

>>>>

Equation \ref{eq:integral-sum_mag} can be rewritten in matrix notation as follows:
\begin{equation}
\mathbf{d}(\mathbf{p}) = \mathbf{A} \mathbf{p} \: ,
\label{eq:predicted-data-vector_mag}
\end{equation}
where $\mathbf{d}(\mathbf{p})$ is the $N \times 1$ predicted data vector with $i$th element defined
be the predicted total-field anomaly $\Delta T_{i}$ (equation \ref{eq:integral-sum_mag}),
$\mathbf{p}$ is the $M \times 1$ parameter vector whose $j$th element is the magnetic moment intensity
$p_{j}$ of the $j$th dipole and $\mathbf{A}$ is the $N \times M$ sensitivity matrix with element 
$ij$ defined by the harmonic function $a_{ij}$ (equation \ref{eq:aij_mag}).
In the classical equivalent-layer technique, the common approach for 
estimating the parameter vector $\mathbf{p}$ from the observed 
total-field anomaly data $\mathbf{d}^{o}$ is solving the least-squares normal equations
\begin{equation}
\mathbf{A}^{\top}\mathbf{A} \: \mathbf{p} = 
\mathbf{A}^{\top} \mathbf{d}^{o} \: .
\label{eq:normal-equations}
\end{equation}
Equation \ref{eq:normal-equations} is usually solved by first computing the Cholesky 
factor $\mathbf{G}$ of matrix $\mathbf{A}^{\top}\mathbf{A}$ and then using it to solve the linear 
systems \citep[][ p. 262]{golub-vanloan2013}:
\begin{equation}
\begin{split}
\mathbf{G} \mathbf{w} &= \mathbf{A}^{\top}\mathbf{d}^{o} \\
\mathbf{G}^{\top} \tilde{\mathbf{p}} &= \mathbf{w}
\end{split} \: ,
\label{eq:classical-method}
\end{equation}
where $\mathbf{w}$ is a dummy variable.
This approach to estimate the parameter vector will be 
referenced throughout this work as the \textit{classical method}.
The computational cost associated with the classical method can be very high
when dealing with large datasets. In the following subsections, we will show how to 
explore the structure of the sensitivity matrix $\mathbf{A}$ and 
efficiently solve the least-squares normal equations (equation \ref{eq:normal-equations}).

@@

Note that the standard CGLS solution (Algorithm \ref{al:std-cgls-algorithm}) requires 
neither inverse matrix nor matrix-matrix product. Instead, it only requires: one matrix-vector 
product out of the loop and two matrix-vector products per iteration (in steps 3 and 6). 
These products can be efficiently computed by using the 2D FFT, as a discrete convolution
(see Appendix A). \citet{takahashi2020convolutional} used this approach
to develop an efficient algorithm for gravity data processing. This modified approach in which
the standard CGLS method is modified to efficiently compute the matrix-vector products will be 
referenced throughout this work as the \textit{convolutional equivalent layer method}.

In this sections we compare the efficiency of the classical (equation \ref{eq:classical-method}), 
standard CGLS (Algorithm \ref{al:std-cgls-algorithm}) and the convolutional equivalent 
layer method (Algorithm \ref{al:std-cgls-algorithm} with matrix-vector products computed 
according to Appendix A). To do this, we compute the total number of 
\textit{flops} associated to them \citep[][ p. 12]{golub-vanloan2013}.

For the classical method, we have $\tfrac{1}{2} N^3$ flops to compute the lower triangle of
$\mathbf{A}^{\top}\mathbf{A}$; $\tfrac{1}{3} N^3$ flops to compute the Cholesky factor
$\mathbf{G}$ of $\mathbf{A}^{\top}\mathbf{A}$ \citep[][ p.~164]{golub-vanloan2013};
$2 \, N^2$ flops to compute the matrix-vector product $\mathbf{A}^{\top} \mathbf{d}^{o}$;
and $2 \, N^2$ flops to solve the triangular systems given by equation \ref{eq:classical-method}
\citep[][ p.~106]{golub-vanloan2013}. The resultant flop count for the classical method is
\begin{equation}
f_{classical} =  \dfrac{5}{6} N^{3} + 4 \, N^{2}\: .
\label{eq:flops-classical-method}
\end{equation}

%======================================================================================
%\subsection{CGLS flops count}
%======================================================================================

For the standard CGLS method (Algorithm \ref{al:std-cgls-algorithm}) we have $2 \, N^2$ to compute
the matrix-vector product $\mathbf{A}^{\top} \mathbf{s}_{(it)}$ out of the loop;
$4 \, N$ in step 1; $2 \, N$ in step 2; $2 \, N^2 + 2 \, N$ in step 3; $2 \, N$ in step 4;
$2 \, N$ in step 5; and $2 \, N^2$ in step 6. The resultant flop count is given by:
\begin{equation}
f_{cgls} =  2 N^{2} + it \, (4 N^{2} + 12 N) \: .
\label{eq:flops-standard-cgls}
\end{equation}

%======================================================================================
%\subsection{Our modified CGLS flops count}
%======================================================================================

To compute the flops count of our method, we need only to replace the flops associated with 
matrix-vector products in the standard CGLS method by those associated with
2D convolution defined in Appendix A, which consists of $\kappa  \, 4 N \log_2(4N)$ flops to
compute the 2D FFT for each matrix $\mathbf{L}_{\boldsymbol{\alpha\beta}}$ (equation 
\ref{eq:L_alpha_beta}); $\kappa  \, 4 N \log_2(4N)$ flops to compute 
$\mathbf{F}_{2Q} \, \mathbf{V} \, \mathbf{F}_{2P}$ via 2D FFT; $24 \, N$ flops to compute the 
Hadamard product $\mathbf{L} \circ \left(\mathbf{F}_{2Q} \, \mathbf{V} \, \mathbf{F}_{2P} \right)$; 
and $\kappa  \, 4 N \log_2(4N)$ flops to compute the IDFT in equation 
\ref{eq:2d-discrete-convolution-complete}. We use $\kappa = 5$ for the \emph{radix-2} algorithm
\citep[][ p.~15]{vanloan1992}. By replacing these flops into Algorithm \ref{al:std-cgls-algorithm},
we obtain the complete number of flops
\begin{equation}
f_{conv} =  \kappa  \, 16 N \log_2(4 N) + 24 N + it \, (\kappa  \, 16 N \log_2 (4 N) + 60 N) \: .
\label{eq:flops-convolutional-method}
\end{equation}

@@

Figure \ref{fig:flops} shows a comparative between the methods varying the number of observation points up to $1,000,000$, where it is possible to observe a reduction of $10^7$ orders of magnitude to estimate parameter vector $\mathbf{\hat{p}}$ in relation to the non-iterative classical method and $10^3$ orders of magnitude in relation to the standart CGLS algorithm using $50$ iterations. 
%A more detailed, step by step, flops count of the classical and CGLS algorithm can be found in Appendix A.

@@

%In Figure \ref{fig:sources_time} a comparison between the time to complete the task to calculate the first column of the BCCB matrix embbeded from the from $\mathbf{A}$ (equation \ref{eq:aij_mag}) by using only one equivalent source, i.e., calculating all six first column of the second derivatives matrices from $\mathbf{H}$ (equation \ref{eq:Hi}) and using four equivalent sources to calculate the four necessary columns from the non-symmetric matrix $\mathbf{A}$ (equation \ref{eq:aij_mag}). Although, very similar in time, with one source a small advantage can be observed as the number of data $N$ increases and goes beyond $N = 200,000$. This test was done from $N = 10,000$ to $N = 700,000$ with increases of $5,625$ observation points.
